Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-01-22 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
sstimer stuff

For example, we can form sums of squares in R using `for` loops, a vectorized
`sum()` or the `crossprod()` function. These have different performance profiles.
Let us form the sums of squares of different length random vectors.

```
 n        t(forloop) : ratio          t(sum) : ratio       t(crossprod)    all.equal 
100           31701  :  2.1996        15541  :  1.0784           14412       TRUE 
1000          35604  :  13.004       2900.1  :  1.0592            2737.9     TRUE 
10000        323969  :  27.826        21051  :  1.8081           11643       TRUE 
1e+05       3120277  :  30.604       291234  :  2.8564          101957       TRUE 
1e+06      31140093  :  29.231      4701221  :  4.4131         1065296       TRUE 
1e+07     317191471  :  21.776     40540123  :  2.7832        14565799       TRUE 
```

These are averages of 100 executions of each sum, and we would expect them to be
stable, but there is some variation, likely due to other processes running in an
active computer, as seen by rerunning the script `sstimer.R`.

```
 n        t(forloop) : ratio          t(sum) : ratio        t(crossprod)   all.equal 
100           35643  :  2.4688        13258  :  0.91831           14437      TRUE 
1000          33172  :  12.142         3689  :  1.3503             2731.9    TRUE 
10000        324269  :  27.958        28028  :  2.4166            11598      TRUE 
1e+05       3132152  :  30.765       304086  :  2.9868           101810      TRUE 
1e+06      31262348  :  28.846      3759337  :  3.4688          1083772      TRUE 
1e+07     313522217  :  21.538     38746311  :  2.6618         14556401      TRUE 
```

We also note the very strange result for `n=1000`. ?? No explanation.
Moreover, results in a Windows 10 virtual machine (under VirtualBox)
give the results

```
 n      t(forloop) : ratio            t(sum) : ratio        t(crossprod)   all.equal 
100         43781  :  2.592126        14442  :  0.8550622         16890      TRUE 
1000        34241  :  12.59323         2452  :  0.9018021          2719      TRUE 
10000      348003  :  19.6058         15038  :  0.8472113         17750      TRUE 
1e+05     3404639  :  20.32475       293930  :  1.75468           67512      TRUE 
1e+06    34379874  :  19.36951      3571310  :  2.012065        1774948      TRUE 
1e+07   376569416  :  17.45253     36871902  :  1.70887        21576778      TRUE 
```

>>>>

For example, we can form sums of squares in R using `for` loops, a vectorized
`sum()` or the `crossprod()` function. These have different performance profiles.
We can time the sums of squares calculation of different length random vectors
using these three approaches and get the ratio to the time using `crossprod()`.
This reveals that the time for summing 100 elements is much longer than that
of summing 1000, which seems very strange. If we first "compile" the functions
using `compiler::cmpfun()` we get the second line of timings, which are ordered
by the size of the vector. Thus we are suspicious that R is trying to compile
the functions "on the fly" and this may be taking time when the program
is initiated. Timings in microseconds were run on a machine with the profile (@RMachID2022)

```
M21:john-Linux-5.13.0-27-generic|Intel(R) Core(TM) i5-10400 CPU @ 2.90GHz|33474072576bytesRAM 
```

```
 n         t(forloop) : ratio         t(sum) : ratio        t(crossprod)  all.equal 
100       26686.85 :  1.883607     12721.49  :  0.8979062   14167.95      TRUE 
           3968.35 :  2.074802       953.36  :  0.4984524    1912.64 
1000      33413.23 :  12.08929      2372.34  :  0.8583399    2763.87      TRUE 
          32894.62 :  11.96442      2319.56  :  0.8436696    2749.37 
10000    325475.7  :  28.14976     15694.36  :  1.357375    11562.29      TRUE 
         331670.1  :  28.24715     16511.91  :  1.40626     11741.72 
1e+05   3234528    :  31.81066    346601.8   :  3.408729   101680.6       TRUE 
        3178637    :  30.75105    346969     :  3.356677   103366.8 
1e+06  31728237    :  29.08711   4057012     :  3.719297  1090801         TRUE 
       31837624    :  29.11575   4051945     :  3.705535  1093485 
```
